---
title: "Investigating samples"
format: 
  html:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
editor: source
---
# Synopsis
Investigating different subsampling procedures for each region, comparing length distributions across surveys, deciding which surveys to remove from models due to data insufficiency.

```{r echo=FALSE,warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(transport)
library(kableExtra)
fish_all <- readRDS("fish_all.rds")
```


```{r echo = FALSE}
s <- fish_all |> 
  # filter(weight_kg > 0, length_cm > 0, !is.na(lon), !is.na(temp), 
  #        year != 2020, sex %in% c("M", "F")) |>
  # mutate(yearc = as.numeric(scale(year, center = T, scale = F)),
  #        tempc = temp - mean(temp),
  #        salinityc = salinity - mean(salinity),
  #        lengthc = log(length_cm) - mean(log(length_cm),na.rm=TRUE)) |>
  # filter(!survey %in% c("northern Bering Sea", "Bering Sea Slope")) |>
  filter(!common_name %in% c("sablefish", "longspine thornyhead")) |>
  #filter(!(common_name == "pacific spiny dogfish" & survey %in% c("Aleutian Islands", "SYN WCHG", "Gulf of Alaska"))) |>
  #filter(!(common_name == "dover sole" & survey == "Aleutian Islands")) |>
  mutate(type = case_when(
    is.na(weight_kg) ~ "length-only",
    !is.na(weight_kg) ~ "length + weight"
  ))|>
  mutate(survey2 = case_when(
    survey == "NWFSC.Combo" ~ "West Coast",
    survey == "SYN HS" ~ "Hecate Strait",
    survey == "SYN QCS" ~ "Queen Charlotte Sound",
    survey == "SYN WCHG" ~ "WC Haida Gwaii",
    survey == "SYN WCVI" ~ "WC Vancouver Island",
    .default = survey
   ))

ss <- s |> filter(weight_kg > 0, length_cm > 0, !is.na(lon), !is.na(temp), 
          year != 2020, sex %in% c("M", "F"))
# fish_all_clean$survey2 <- factor(fish_all_clean$survey2, levels = c("Aleutian Islands", "Bering Sea Slope", "northern Bering Sea", "eastern Bering Sea", "Gulf of Alaska", "Hecate Strait", "Queen Charlotte Sound", "WC Haida Gwaii", "WC Vancouver Island","West Coast"))
```

# Length distributions
## Histogram by survey
Do the length distributions of the length-only and length+weight data entries match? Where are areas with sparse data?

- NWFSC only has observations with length+weight. Data is from the surveyjoin-db age_and_length.rds
- AFSC mostly has observations with length+weight. Data is from surveyjoin-db afsc_specimen.rds
- PBS has both types. Data is from fishyplots pbs-gfdata.rds

Note: there are 23,895 rows removed from PBS due to missing length information, true for all following plots.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(s, aes(x = length_cm, fill = type)) +
  geom_histogram() +
  facet_wrap(~survey) +
  scale_fill_brewer(palette= "Dark2")+
  theme_minimal()
```
## CDFs by survey

Comparing CDFs (cumulative distribution functions) of length-only with length+weight fish entries, we see that

- Alaska looks weird since there are relatively few length-only observations
- The West Coast random subsampling is reflected
- Canada maybe is slightly oversampling larger fish
- Generally the length+weight CDFs for each region look pretty similar
```{r echo=FALSE,warning=FALSE}
s |> filter(!survey %in% c("northern Bering Sea", "Bering Sea Slope")) |>
ggplot(aes(x = length_cm, color = type)) +
  stat_ecdf() +
  facet_wrap(~survey)+
  scale_color_brewer(palette = "Dark2")+
  theme_minimal()
```
## Density plots by survey
Surveys with insufficient data are apparent here (eg. dogfish in Aleutians). Some surveys/species have notably bimodal-ish distributions.

May need to look into depth gradients for each region. NWFSC is known for deeper depths, which may lead to sampling longer/bigger fish. There may be an example of this with shortspine thornyhead below.
```{r echo=FALSE,warning=FALSE}
# s |> filter(!survey %in% c("northern Bering Sea", "Bering Sea Slope")) |>
# ggplot(aes(x = length_cm, color = type)) +
#   geom_density() +
#   facet_wrap(~survey2)+
#   scale_color_brewer(palette = "Dark2")+
#   theme_minimal()

s |> filter(!survey %in% c("northern Bering Sea", "Bering Sea Slope"), !common_name %in% c("longspine thornyhead","sablefish")) |>
ggplot(aes(x = length_cm, color = survey)) +
  labs(color="survey")+
  geom_density() +
  facet_wrap(~common_name)+
  scale_color_brewer(palette = "Dark2")+
  theme_minimal()
```

### Depth gradients across regions
Looks like NWFSC definitely does have much more of a right skew towards deeper depths, although it varies by species. Perhaps it is most notable for Shortspine thornyhead, where we also see a bit of right skew in the graph above.
```{r echo=FALSE, warning=FALSE}
s |> filter(!survey %in% c("northern Bering Sea", "Bering Sea Slope"), !common_name %in% c("longspine thornyhead","sablefish")) |>
ggplot(aes(x = depth, color = region)) +
  labs(color="survey")+
  geom_density() +
  facet_wrap(~common_name)+
  scale_color_brewer(palette = "Dark2")+
  theme_minimal()
```

## Distribution tests and CDF by species
Wasserstein test: determine if two samples come from the same distribution. Quantifies the amount of transport you would need to get from one distribution to the other.

Arrowtooth flounder, dover sole, and pacific ocean perch look pretty good.
Dogfish and shortspine look less great. Dogfish in Alaska are very few samples.
```{r echo =FALSE,warning=FALSE}
s |>
  filter(type == "length + weight") |>
ggplot(aes(x = length_cm, color = region)) +
  stat_ecdf() +
  facet_wrap(~common_name)+
  scale_color_brewer(palette = "Dark2")+
  theme_minimal()

for (i in list("arrowtooth flounder", "dover sole", "pacific spiny dogfish", "shortspine thornyhead", "pacific ocean perch")) {
  ca1 <- ss |> filter(region == "pbs", common_name == i)|>pull(length_cm)
  ca2 <- ss |> filter(region == "nwfsc",common_name == i) |>pull(length_cm)
  ca3 <- ss |> filter(region == "afsc", common_name == i) |> pull(length_cm) 
  message("Species: ", i, ". PBS sd: ", round(sd(ca1),2), ". AFSC sd: ", round(sd(ca3),2), ". NWFSC sd: ", round(sd(ca2),2))  
  message("PBS and NWFSC: ", round(wasserstein1d(ca1,ca2,p=1),2), " cm.")
  message("PBS and AFSC: ", round(wasserstein1d(ca1,ca3,p=1),2), " cm.")
  message("NWFSC and AFSC: ", round(wasserstein1d(ca2,ca3,p=1),2), " cm.")
  message("")
}
```

# Tables by year and survey - removing data insufficient surveys

## Arrowtooth flounder
- Remove northern Bering Sea (only 1 year)
- Remove Bering Sea Slope (only 3 years)
```{r echo=FALSE}
ss |> 
  filter(common_name == "arrowtooth flounder") |>
  count(common_name,survey,year) |>kbl()
```
## Dover sole
- Remove Aleutian Islands (only 1 year)
- Remove all Bering Sea (no data)
```{r echo=FALSE}
ss |> 
  filter(common_name == "dover sole") |>
  count(common_name,survey,year) |>kbl()
```

## Shortspine thornyead
- Remove Bering Sea Slope (only 1 year)
- Remove all other Bering Sea (no data)

- Hecate Strait no data before 2011...
- Queen Charlotte Sound no data before 2009...
- Haida Gwaii and Vancouver Island no data before 2010...
```{r echo=FALSE}
ss |> 
  filter(common_name == "shortspine thornyhead") |>
  count(common_name,survey,year) |>kbl()
```

## Pacific ocean perch
- Remove Bering Sea Slope (no data after 2016)
- Remove all other Bering Sea (no data)
```{r echo=FALSE}
ss |> 
  filter(common_name == "pacific ocean perch") |>
  count(common_name,survey,year) |>kbl()
```

## Pacific spiny dogfish
- Remove Aleutians (only 1 year)
- Remove Gulf of Alaska (only 4 years of substantial data, and none after 2011)
- Remove Haida Gwaii (too few observations)
- Hecate Strait no substantial data before 2009...
```{r echo=FALSE}
ss |> 
  filter(common_name == "pacific spiny dogfish") |>
  count(common_name,survey,year) |>kbl()
```



