---
title: "all_species"
format: html
editor: source
---
Notes:
- time series for each survey for st effect, averg across year for each survey domain
- Experiment with diffusion
- Add trawl id as random effect.
- Put date/time as column in aic table

# Imports
```{r warning = FALSE, message = FALSE}
library(sdmTMB)
library(readr)
library(ggplot2)
library(surveyjoin)
library(stats)
library(rnaturalearth)
library(stringr)
library(dplyr)
library(lubridate)
library(broom)
library(tidyr)
library(sf)
library(patchwork)
library(PNWColors)
library(RhpcBLASctl)
RhpcBLASctl::blas_set_num_threads(1)
RhpcBLASctl::omp_set_num_threads(1)
options(scipen = 999)

RhpcBLASctl::blas_get_num_procs()
RhpcBLASctl::omp_get_max_threads()
```

# Data cleaning
```{r}
# Biological data
fish_all <- readRDS("fish_all.rds")

# Filters: no 2020 PBS fish, no unsexed fish, no entries with missing coordinates, no zeroed lengths/weights that will crash the models, no northern Bering or Bering slope, no dogfish in Alaska, no sole in Aleutians
fish_all_clean <- fish_all |> 
  filter(weight_kg > 0, length_cm > 0, !is.na(lon), !is.na(temp), 
         year != 2020, sex %in% c("M", "F")) |>
  filter(!survey %in% c("northern Bering Sea", "Bering Sea Slope")) |>
  filter(!common_name %in% c("sablefish", "longspine thornyhead")) |>
  filter(!(common_name == "pacific spiny dogfish" & survey %in% c("Aleutian Islands", "SYN WCHG", "Gulf of Alaska"))) |>
  filter(!(common_name == "dover sole" & survey == "Aleutian Islands"))
rm(fish_all)
```

# Comparing 4 model structures
```{r}
fsh_all_structures <- function (species, cutoff) {
  # Preprocessing --> removing entries beyond -3/3 SDs from the mean residual
  a <- fish_all_clean |> 
    filter(common_name == species)
  a1 <- lm(log(weight_kg) ~ log(length_cm), data = a)
  a <- a |>
    mutate(residuals = residuals(a1)) |>
    mutate(over_3sd = abs(residuals - mean(residuals)) > 3*sd(residuals))
  message("Residuals > 3 SD from mean: ", sum(a$over_3sd))
  a <- a|> filter(over_3sd == FALSE)
  
  # Make mesh
  a <- sdmTMB::add_utm_columns(a, ll_names = c("lon", "lat"), utm_crs = 3338)
  mesh <- sdmTMB::make_mesh(a, xy_cols = c("X","Y"), cutoff = cutoff)
  message("Mesh: ", mesh$mesh$n, " Modeling ", species, " A...")
  
  # Modeling
  A <- try(sdmTMB(log(weight_kg) ~ log(length_cm) + as.factor(year) + as.factor(sex),
            mesh = mesh,
            spatial = "on",
            time = "year",
            spatiotemporal = "rw",
            anisotropy = TRUE,
            data = a))
  
  message("B...")
  # Does the l-w relationship vary spatially?
  B <- try(sdmTMB(log(weight_kg) ~ log(length_cm)*as.factor(region) + as.factor(year) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log(length_cm),
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a))
  message("C...")
  # Are there region-specific year trends in condition? (simple approach)
  C <- try(sdmTMB(log(weight_kg) ~ log(length_cm) + as.factor(year)*as.factor(region) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a))
  message("D...")
  # Is condition better predicted when year has spatial field variance or as an interaction in the fixed effects (C)? ----- NO CONVERGENCE, even with using residuals
  D <- try(sdmTMB(log(weight_kg) ~ log(length_cm) + yearc + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ yearc,
              #time = "year",
              #spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a))
  
  return(list(A,B,C,D))
}

# Comparing structures --> model B preferred
AIC(arr[[1]],arr[[2]],arr[[3]],arr[[4]]) |> arrange(AIC)
AIC(shortspine[[1]], shortspine[[2]]) |> arrange(AIC)
AIC(dover[[1]],dover[[2]]) |> arrange(AIC)
AIC(dogfish[[1]], B) |> arrange(AIC)
AIC(perch[[1]], perch[[2]], perch[[3]],perch[[4]]) |> arrange(AIC)
```

# Base model (B)
```{r}
fsh_b <- function(species, cutoff, center = TRUE, dist = c("gaussian", "student")) {
  dist <- match.arg(dist)
  fam <- switch(dist, 
                gaussian = gaussian(link = "identity"),
                student = student(link = "identity", df = NULL))
  
  # Center variables (species-specific)
  a <- fish_all_clean |> 
    filter(common_name == species) |>
    mutate(yearc = as.numeric(scale(year, center = T, scale = F)),
           tempc = temp - mean(temp),
           salinityc = salinity - mean(salinity),
           log_length = log(length_cm),
           log_lengthc = log_length - mean(log_length))
  
  # Preprocessing --> removing entries beyond -3/3 SDs from the mean residual
  a1 <- lm(log(weight_kg) ~ log(length_cm), data = a)
  a <- a |>
    mutate(residuals = residuals(a1)) |>
    mutate(over_3sd = abs(residuals - mean(residuals)) > 3*sd(residuals))
  message("Residuals > 3 SD from mean: ", sum(a$over_3sd), " (", round(sum(a$over_3sd)/nrow(a),3), "%)")
  a <- a|> filter(over_3sd == FALSE)
  
  # Make mesh
  a <- sdmTMB::add_utm_columns(a, ll_names = c("lon", "lat"), utm_crs = 3338)
  mesh <- sdmTMB::make_mesh(a, xy_cols = c("X","Y"), cutoff = cutoff)
  message("Mesh: ", mesh$mesh$n, " Modeling ", species, " B...")
  
  # Modeling structure B.
  if (center == FALSE) {
  B <- try(sdmTMB(log(weight_kg) ~ log(length_cm)*as.factor(region) + as.factor(year) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log(length_cm),
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  }
  if (center == TRUE) {
    B <- try(sdmTMB(log(weight_kg) ~ log_lengthc*as.factor(region) + as.factor(year) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log_lengthc,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  }

  return(B)
}

############################ 
# Dogfish --> change spatiotemporal to iid
a <- fish_all_clean |> filter(common_name == "pacific spiny dogfish") |> mutate(yearc = as.numeric(scale(year, center = T, scale = F)),
           tempc = temp - mean(temp), salinityc = salinity - mean(salinity), log_length = log(length_cm),
           log_lengthc = log_length - mean(log_length))
a1 <- lm(log(weight_kg) ~ log(length_cm), data = a)
a <- a |> mutate(residuals = residuals(a1)) |>
  mutate(over_3sd = abs(residuals - mean(residuals)) > 3*sd(residuals))
message("Residuals > 3 SD from mean: ", sum(a$over_3sd), " (", round(sum(a$over_3sd)/nrow(a),3), "%)")
a <- a|> filter(over_3sd == FALSE)
a <- sdmTMB::add_utm_columns(a, ll_names = c("lon", "lat"), utm_crs = 3338)
mesh <- sdmTMB::make_mesh(a, xy_cols = c("X","Y"), cutoff = 35)
dogfish <- sdmTMB(log(weight_kg) ~ log_lengthc*as.factor(region) + as.factor(year) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log_lengthc,
              #time = "year",
              #spatiotemporal = "iid", 
              anisotropy = TRUE,
              data = a)
############################

# Run models
# arrowtooth <- fsh_b("arrowtooth flounder", 100, dist = "student")
# dover <-  fsh_b("dover sole", cutoff = 74) 
# #dogfish <- fsh_b("pacific spiny dogfish", cutoff = 35) 
# shortspine <- fsh_b("shortspine thornyhead", cutoff = 86, dist = "student")
# perch <- fsh_b("pacific ocean perch", cutoff = 75)

# Save models
saveRDS(arrowtooth, "models/arrowtooth.rds")
saveRDS(dover,"models/dover.rds")
saveRDS(dogfish,"models/dogfish.rds")
saveRDS(shortspine,"models/shortspine.rds")
saveRDS(perch,"models/perch.rds")

# Read models
arrowtooth <- readRDS("models/arrowtooth.rds")
dogfish <- readRDS("models/dogfish.rds")
dover <- readRDS("models/dover.rds")
perch <- readRDS("models/perch.rds")
shortspine <- readRDS("models/shortspine.rds")
```

# Base model - Predictions and mapping
Arrowtooth -- remove northern Bering Sea, Bering Sea Slope
Dover -- remove all Bering Sea, Aleutians
Shortspine -- remove all Bering Sea
Dogfish -- remove all Alaska, Haida Gwaii
Perch -- remove all Bering Sea, remove southern California (lat < 40)
```{r}
# Get base maps
states <- ne_download(scale = 50, type = "states", category = "cultural", returnclass = "sf")
states <- states |> filter(admin == "United States of America")
canada <- ne_download(scale = 50, type = "states", category = "cultural", returnclass = "sf")
canada <- canada |> filter(admin == "Canada") |> select(name, geometry)
mexico <- ne_countries(scale = "medium", returnclass = "sf", country = "mexico") |> 
  select(name = admin, geometry)
alaska <- states |> filter(.data$name == "Alaska")

# Combine, get equal areas crs, crop
proj <- bind_rows(states, canada, mexico, alaska) |>
  st_transform(4326) |>
  st_crop(c(xmin = -170, ymin = 25, xmax = -115, ymax = 65)) |>
  st_transform(3338)
rm(states, canada, mexico, alaska)
gc()
```

```{r}
predictr <- function(model) {
  # Get grids from surveyjoin and combine, removing species-specific data-sparse areas
  grid <- bind_rows(surveyjoin::nwfsc_grid, surveyjoin::dfo_synoptic_grid, surveyjoin::afsc_grid) |>
    filter(!survey %in% c("Northern Bering Sea Crab/Groundfish Survey - Eastern Bering Sea Shelf Survey Extension", "Eastern Bering Sea Slope Bottom Trawl Survey", "Triennial"))
  grid <- add_utm_columns(grid, ll_names = c("lon", "lat"), utm_crs = 3338)
  
  if (unique(model$data$common_name) == "pacific spiny dogfish") {
    grid <- grid |> filter(!grepl("Bering", survey), !grepl("Aleutian",survey), !grepl("Gulf",survey), survey!="SYN WCHG")}
  
  if (unique(model$data$common_name) == "dover sole") {
    grid <- grid |> filter(!grepl("Aleutian", survey))}
  
  if (unique(model$data$common_name) %in% c("dover sole","shortspine thornyhead","pacific ocean perch", "pacific spiny dogfish")) {
    grid <- grid |> filter(!grepl("Eastern Bering Sea Crab", survey))}
  
  if (unique(model$data$common_name) %in% c("pacific ocean perch")) {
    grid <- grid |> filter(lat >= 40)}
  
  if (unique(model$data$common_name) == "arrowtooth flounder") {
    grid <- grid |> filter(lat >= 34)
  }
  
  # Add time, length, sex. Has to be after the filters!!
  grid <- expand_grid(grid, year = unique(model$data$year)) |>
    mutate(length_cm = mean(model$data$length_cm, na.rm = T)) |>
    mutate(log_lengthc = mean(model$data$log_lengthc, na.rm=T)) |>
    mutate(sex = "F") |>
    mutate(region = case_when(
      survey == "NWFSC.Combo" ~ "nwfsc",
      survey %in% c("SYN QCS", "SYN HS", "SYN WCVI", "SYN WCHG") ~ "pbs",
      .default = "afsc"))
  
  pred <- predict(model, grid)
  pred <- pred |>
    mutate(species = unique(model$data$common_name)) |>
    st_as_sf(coords = c("lon", "lat"), crs = 4326) |>
    st_transform(3338) |>
    mutate(x = st_coordinates(geometry)[,1],
           y = st_coordinates(geometry)[,2]) |>
    st_drop_geometry()
  rm(grid)
  gc()
}

arrowtooth_pred <- predictr(arrowtooth$base_centered)
dover_pred <- predictr(dover)
dogfish_pred <- predictr(dogfish)
perch_pred <- predictr(perch)
shortspine_pred <- predictr(shortspine)

mapper <- function(species, preds, fill) {
  
  if (fill == "zeta_s" & !any(grepl("zeta_s_log_",names(preds)))) {
    fill <- "zeta_s_log(length_cm)"
  } else if (fill == "zeta_s" & any(grepl("zeta_s_log_",names(preds)))) {
    fill <- "zeta_s_log_lengthc"
  }
  
  r <- max(abs(preds[[fill]]))
  
  g <- ggplot() +
    stat_summary_hex(data = preds, aes(x = x, y = y, z = .data[[fill]]), bins = 50) +
    scale_fill_gradientn(name = fill,
                         colours = rev(pnw_palette("Shuksan2")),
                         limits = c(-r,r)
                         ) +
    geom_sf(data = proj) +
    geom_sf_text(data = proj, aes(label = name), size = 2, check_overlap = F) +
    theme_minimal() +
    labs(x = "", y = "", title = species, color = "log(length)")
return(g)
}

j <- mapper("Arrowtooth flounder", arrowtooth_pred, "omega_s") +
  mapper("Arrowtooth flounder", arrowtooth_pred, "zeta_s") + 
  mapper("Dover sole", dover_pred, "omega_s") +
  mapper("Dover sole", dover_pred, "zeta_s") + 
  mapper("Shortspine thornyhead",shortspine_pred, "omega_s") +
  mapper("Shortspine thornyhead",shortspine_pred, "zeta_s") +
  mapper("Pacific spiny dogfish", dogfish_pred, "omega_s") +
  mapper("Pacific spiny dogfish", dogfish_pred, "zeta_s") + 
  mapper("Pacific ocean perch", perch_pred, "omega_s") +
  mapper("Pacific ocean perch", perch_pred, "zeta_s") + 
  patchwork::plot_annotation(title = "Log(length) variance of female fish",
                             theme = theme(plot.title = element_text(size = 28)))+
  plot_layout(ncol=2) 

ggsave("figures/test_spatial_trends.png", j, width = 12, height = 18)
```


# Environmental covariates
```{r}
check_aic <- function(model_list, aic_table = NULL, model_list_name = NULL) {

  # If not being called within the fsh() then dynamically name outgoing object
  if (is.null(model_list_name)) {
    model_list_name <- deparse(substitute(model_list))
  }
  
  # Loop through models in list and perform sanity check + AIC calculation
  results <- lapply(names(model_list), function(nm) {
    mod <- model_list[[nm]]

    if (is.null(mod) || !inherits(mod, "sdmTMB")) {
      message(sprintf("Skipping '%s': not a valid sdmTMB object", nm))
      return(data.frame(model = nm, aic = NA_real_, converged = NA,
                        sanity_ok = NA, note = "not a valid model object",
                        formula = NA))
    }

    
    sanity_res <- suppressWarnings(suppressMessages(sanity(mod, silent = TRUE)))
      
    #small_sigma <- any(grepl("smaller than 0.01", sanity_messages))
    #sanity_ok <- isTRUE(sanity_res$all_ok)
    
    sigma_Z_val <- tryCatch(
    tidy(mod, effects = "ran_pars") |> filter(term == "sigma_Z") |> pull(estimate),
    error = function(e) NA_real_)
    small_sigma <- length(sigma_Z_val) > 0 && !is.na(sigma_Z_val) && sigma_Z_val < 0.01
      
    sanity_ok <- (isTRUE(sanity_res$hessian_ok)) &&
      (isTRUE(sanity_res$eigen_values_ok)) &&
      (isTRUE(sanity_res$nlminb_ok)) &&
      (isTRUE(sanity_res$range_ok)) &&
      (isTRUE(sanity_res$gradients_ok)) &&
      (isTRUE(sanity_res$se_magnitude_ok)) &&
      (isTRUE(sanity_res$se_na_ok)) &&
      (isTRUE(sanity_res$sigmas_ok) || small_sigma)
    
    if (!sanity_ok) {
      failed_checks <- names(sanity_res)[vapply(sanity_res, isFALSE, logical(1))]
      message(sprintf("Skipping '%s': failed sanity checks: %s",
                       nm, paste(failed_checks, collapse = ", ")))
      return(data.frame(model = nm, aic = NA_real_, converged = NA,
                         sanity_ok = FALSE,
                         note = paste(failed_checks, collapse = "; "),
                        formula = as.character(mod$formula)))
    }

    aic_val <- tryCatch(
      AIC(mod),
      error = function(e) NA_real_,
      warning = function(w) NA_real_
    )
    failed_checks <- names(sanity_res)[vapply(sanity_res, isFALSE, logical(1))]
    data.frame(model = nm, aic = aic_val, converged = TRUE,
               sanity_ok = TRUE, note = paste(paste(failed_checks, collapse=", "), "Small sigma: ", small_sigma),
               formula = as.character(mod$formula))
  })

  # Attach all results rows in order of descending AIC
  out <- do.call(bind_rows, results)
  out <- out[order(out$aic), ]
  rownames(out) <- NULL
  
  # If there is an existing table, attach that and order by descending AIC
  if (!is.null(aic_table)) {
    aic_table <- aic_table |> filter(!model %in% c(out$model))
    out <- bind_rows(aic_table, out)
    out <- out[order(out$aic), ]
  }
  write_csv(out, paste0("models/", model_list_name, "_aic.csv")) 
}
```

```{r}

fsh_env <- function(species, obj_name, vars = c(NULL, "base", "base_re", "base_centered", "temp", "salinity", "temp_salinity", "year_removal", "st_removal", "temp_region", "temp_region2", "spline_removal", "temp_svc", "diffusion1", "diffusion2"), center = FALSE, dist = c("gaussian", "student"), save_models = FALSE) {
  
  if (length(vars) > 0) {
    # Dynamically update distribution
    dist <- match.arg(dist)
    fam <- switch(dist, 
                  gaussian = gaussian(link = "identity"),
                  student = student(link = "identity", df = NULL))
    
    # Dynamically update cutoff value
    cutoff <- switch(species,
                     "arrowtooth flounder" = 100,
                     "dover sole" = 74,
                     "pacific spiny dogfish" = 35,
                     "shortspine thornyhead" = 86,
                     "pacific ocean perch" = 75)
    
    # Center variables (species-specific)
    a <- fish_all_clean |> 
      filter(common_name == species) |>
      mutate(yearc = as.numeric(scale(year, center = T, scale = F)),
             tempc = temp - mean(temp),
             salinityc = salinity - mean(salinity),
             log_length = log(length_cm),
             log_lengthc = log_length - mean(log_length))
    
    if (center == TRUE) {
      a$length_var <- a$log_lengthc
    } else {
      a$length_var <- a$log_length
    }
    
    # Preprocessing --> removing entries beyond -3/3 SDs from the mean residual
    a1 <- lm(log(weight_kg) ~ log(length_cm), data = a)
    a <- a |>
      mutate(residuals = residuals(a1)) |>
      mutate(over_3sd = abs(residuals - mean(residuals)) > 3*sd(residuals))
    message("Residuals > 3 SD from mean: ", sum(a$over_3sd), " (", round(sum(a$over_3sd)/nrow(a),3), "%)")
    a <- a|> filter(over_3sd == FALSE)
    
    # Make mesh
    if ("diffusion2" %in% vars) {
      missing_year <- data.frame(year = 2020, lon = -179, lat = 70)
      a <- a |> bind_rows(missing_year)
    }
    
    a <- sdmTMB::add_utm_columns(a, ll_names = c("lon", "lat"), utm_crs = 3338)
    mesh <- sdmTMB::make_mesh(a, xy_cols = c("X","Y"), cutoff = cutoff)
    message("Mesh: ", mesh$mesh$n, ".")
  }
  
  # Create empty results list
  results <- list()
  
  # base
  if ("base" %in% vars) {
    message("Modeling ", species, " base...")
    results$base <- try(sdmTMB(log(weight_kg) ~ log_length*as.factor(region) + as.factor(year) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log_length,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  }
  
  if ("base_re" %in% vars) {
    message("Modeling ", species, " base...")
    results$base_re <- try(sdmTMB(log(weight_kg) ~ log_length*as.factor(region) + as.factor(year) + as.factor(sex) + (1|trawl_id),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log_length,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  }
  
  
  if ("base_centered" %in% vars) {
    message("Modeling ", species, " base (centered log_length)...")
    results$base_centered <- try(sdmTMB(log(weight_kg) ~ log_lengthc*as.factor(region) + as.factor(year) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log_lengthc,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  }
  
  # add temperature
  if ("temp" %in% vars) {
  message("Modeling ", species, " temperature only...")
  results$temp <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex) + s(tempc,k=4),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ length_var,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  } 
  
  # add salinity
  if ("salinity" %in% vars) {
  message("Modeling ", species, " salinity only...")
  results$salinity <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex)+
                     s(salinityc,k=4),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ length_var,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  } 
  
  # both
  if ("temp_salinity" %in% vars) {
  message("Modeling ", species, " temperature and salinity...")
  results$temp_salinity <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex) + s(tempc,k=4)+ s(salinityc,k=4),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ length_var,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  } 
  
  # Year removal
  if ("year_removal" %in% vars) {
  message("Modeling ", species, " year removal...")
  if (species %in% c("pacific ocean perch")) {
    formula <- log(weight_kg) ~ length_var*as.factor(region) + as.factor(sex) + s(tempc,k=4)
  } else {
    formula <- log(weight_kg) ~ length_var*as.factor(region) + as.factor(sex) + s(tempc,k=4) + s(salinityc,k=4)
  }
  
  results$year_removal <- try(sdmTMB(formula = formula,
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ length_var,
              time = "year",
              spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  }
  
  # Spatiotemporal removal
  if ("st_removal" %in% vars) {
  message("Modeling ", species, " spatiotemporal removal...")
  if (species %in% c("pacific ocean perch")) {
    formula <- log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex) + s(tempc,k=4)
  } else {
    formula <- log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex) + s(tempc,k=4) + s(salinityc,k=4)
  }
  
  results$st_removal <- try(sdmTMB(formula = formula,
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ length_var,
              #time = "year",
              #spatiotemporal = "rw",
              anisotropy = TRUE,
              data = a,
              family = fam))
  } 
  
  if ("temp_region" %in% vars) {
    message("Modeling ", species, " with temperature:region interaction...")
    a$region <- as.factor(a$region)
    results$temp_region <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex) + s(tempc,k=4, by =region),
                            mesh = mesh,
                            spatial = "on",
                            spatial_varying = ~ length_var,
                            time = "year",
                            spatiotemporal= "rw",
                            anisotropy = TRUE,
                            data = a,
              family = fam))
  }
  
  if ("temp_region2" %in% vars) {
    message("Modeling ", species, " with temperature:region interaction with year removal...")
    a$region <- as.factor(a$region)
    results$temp_region2 <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(sex) + s(tempc,k=4, by =region),
                            mesh = mesh,
                            spatial = "on",
                            spatial_varying = ~ length_var,
                            time = "year",
                            spatiotemporal= "rw",
                            anisotropy = TRUE,
                            data = a,
              family = fam))
  }
  
  if ("spline_removal" %in% vars) {
    message("Modeling ", species, " removing splines...")
    results$spline_removal <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex) + salinityc,
                            mesh = mesh,
                            spatial = "on",
                            spatial_varying = ~ length_var,
                            time = "year",
                            spatiotemporal= "rw",
                            anisotropy = TRUE,
                            data = a,
              family = fam))
  } 
  
  if ("temp_svc" %in% vars) {
    message("Modeling ", species, " temp as SVC...")
    results$temp_svc <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex) + tempc,
                            mesh = mesh,
                            spatial = "on",
                            spatial_varying = ~ length_var + tempc,
                            time = "year",
                            spatiotemporal= "rw",
                            anisotropy = TRUE,
                            data = a,
              family = fam))
  } 
  
  if ("diffusion1" %in% vars) {
    message("Modeling ", species, " temp as spatial diffusion process...")
    results$diffusion1 <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex),
                            mesh = mesh,
                            spatial = "on",
                            spatial_varying = ~ length_var,
                            nonlocal_formula = ~ diffusion(tempc),
                            time = "year",
                            spatiotemporal= "rw",
                            anisotropy = TRUE,
                            data = a,
              family = fam))
  }
  
  if ("diffusion2" %in% vars) {
    message("Modeling ", species, " temp as spatial and temporal diffusion process...")
    
    results$diffusion2 <- try(sdmTMB(log(weight_kg) ~ length_var*as.factor(region) + as.factor(year) + as.factor(sex),
                            mesh = mesh,
                            spatial = "on",
                            spatial_varying = ~ length_var,
                            nonlocal_formula = ~ time_lag(tempc),
                            time = "year",
                            spatiotemporal= "rw",
                            anisotropy = TRUE,
                            data = a,
              family = fam))
  }
  
  # Getting the model object from env/files if no models are requested - only generate AIC table from existing list
  if (length(vars) == 0 || length(results) == 0) {
    message("No models called. Pulling existing models instead from environment...")
    if (exists(obj_name, envir = .GlobalEnv, inherits = FALSE)) {
         results <- get(obj_name, envir = .GlobalEnv)
       } else {
         message("No existing models found in environment. Checking files...")
         if (file.exists(paste0("models/", obj_name, ".rds"))) {
           results <- readRDS(paste0("models/", obj_name, ".rds"))
         } else {
           message("No models found in files.")
           return(invisible(NULL))
         }
       }
  } else if (length(vars) > 0) {
    rm(a, a1, mesh)
    gc()
  }
  
  # Construct AIC tables. Pull existing tables and overwrite if necessary.
  message("Performing sanity check and constructing AIC table...")
  existing_csv <- paste0("models/", obj_name, "_aic.csv")
  prior_table <- if (file.exists(existing_csv)) read_csv(existing_csv, show_col_types = FALSE) else NULL
  check_aic(results, aic_table = prior_table, model_list_name = obj_name)
  
  # Save model list object if requested
  if (save_models == TRUE) {
    message("Saving models...")
    
    existing_saved <- if (file.exists(paste0("models/", obj_name, ".rds"))) {
      readRDS(paste0("models/", obj_name, ".rds"))
    } else {
      list()
    }
    existing_saved[names(results)] <- results
    assign(obj_name, existing_saved, envir = .GlobalEnv)
    saveRDS(existing_saved, paste0("models/", obj_name, ".rds"))
  }
  rm(results)
  invisible(gc())
}

############################ 
# Dogfish --> no convergence
a <- fish_all_clean |> filter(common_name == "pacific spiny dogfish")
a1 <- lm(log(weight_kg) ~ log(length_cm), data = a)
a <- a |>
  mutate(residuals = residuals(a1)) |>
  mutate(over_3sd = abs(residuals - mean(residuals)) > 3*sd(residuals))
message("Residuals > 3 SD from mean: ", sum(a$over_3sd))
a <- a|> filter(over_3sd == FALSE)|>
  mutate(tempc = temp - mean(temp),
        salinityc = salinity - mean(salinity),
        log_length = log(length_cm),
        log_lengthc = log_length - mean(log_length))

a <- sdmTMB::add_utm_columns(a, ll_names = c("lon", "lat"), utm_crs = 3338)
mesh <- sdmTMB::make_mesh(a, xy_cols = c("X","Y"), cutoff = 35)
dogfish <- list()
dogfish$base_centered <- sdmTMB(log(weight_kg) ~ log_lengthc*as.factor(region) + as.factor(year) + as.factor(sex),
              mesh = mesh,
              spatial = "on",
              spatial_varying = ~ log_lengthc,
              time = "year",
              #spatiotemporal = "iid", 
              anisotropy = TRUE,
              data = a)
rm(a, a1, mesh)
gc()
saveRDS(dogfish, "models/dogfish.rds")
############################


# Creating AIC tables for candidate model comparisons
check_sanity <- function(species_env) {
  invisible(lapply(names(species_env), function(nm) {
    cat("\n---", nm, "---\n")
    sanity(species_env[[nm]])
  }))
}

# Saving models for further analysis and visualization
fsh_env("arrowtooth flounder", "arrowtooth", vars = c("base_centered", "year_removal"), dist = "student", save_models = TRUE)
fsh_env("dover sole", "dover", vars = c("base_centered", "year_removal"), save_models = TRUE)
fsh_env("shortspine thornyhead", "shortspine", vars = c("base_centered", "year_removal"), dist = "student", save_models = TRUE)
fsh_env("pacific ocean perch", "perch", vars = c("base_centered", "year_removal"), save_models = TRUE)


# Arrowtooth -  year removal preferred (based on AIC temp+salinity model). No convergence for just salinity or st_removal, or temp*region interaction, convergence when spline removed for salinity.
# Centered: Year removal preferred (temp+salinity). No convergence for temp*region.

fsh_env("arrowtooth flounder", obj_name = "arrowtooth", center = FALSE, dist = "student", vars = c("base_re"))

# Dover - all converge. year removal (temp+salinity) and temp*region preferred.
# Centered - all converge. year removal (temp+salinity) and temp*region preferred.

fsh_env("dover sole", obj_name = "dover", center = FALSE, vars = c("base_re"))
fsh_env("dover sole", obj_name = "dover", vars = NULL)


# Shortspine - temp+salinity and year removal (AIC and BIC). temp*region non-convergence. 
# Centered - temp only

fsh_env("shortspine thornyhead", obj_name = "shortspine", center = FALSE, vars = c("base_re"))

# Perch - temp*region preferred. temp, salinity both below base (because uncentered).
# Centered - temp*region converged and preferred.

fsh_env("pacific ocean perch", obj_name = "perch", center = FALSE, vars = c("base_re"))

fsh_env("pacific ocean perch", obj_name = "perch", vars = NULL)

```


```{r}
modeler <- function(species, obj_name, vars = c(NULL, "base", "base_re", "base_centered", "temp", "salinity", "temp_salinity", "year_removal", "st_removal", "temp_region", "temp_region2", "spline_removal", "temp_svc", "diffusion1", "diffusion2"), center = FALSE, dist = c("gaussian", "student"), save_models = FALSE) {
  
  if (length(vars) > 0) {
    # Dynamically update distribution
    dist <- match.arg(dist)
    fam <- switch(dist, 
                  gaussian = gaussian(link = "identity"),
                  student = student(link = "identity", df = NULL))
    
    # Dynamically update cutoff value
    cutoff <- switch(species,
                     "arrowtooth flounder" = 100,
                     "dover sole" = 74,
                     "pacific spiny dogfish" = 35,
                     "shortspine thornyhead" = 86,
                     "pacific ocean perch" = 75)
    
    # Center variables (species-specific)
    a <- fish_all_clean |> 
      filter(common_name == species) |>
      mutate(yearc = as.numeric(scale(year, center = T, scale = F)),
             tempc = temp - mean(temp),
             salinityc = salinity - mean(salinity),
             log_length = log(length_cm),
             log_lengthc = log_length - mean(log_length))
    
    # Preprocessing --> removing entries beyond -3/3 SDs from the mean residual
    a1 <- lm(log(weight_kg) ~ log(length_cm), data = a)
    a <- a |> mutate(residuals = residuals(a1)) |>
      mutate(over_3sd = abs(residuals - mean(residuals)) > 3*sd(residuals))
    message("Residuals > 3 SD from mean: ", sum(a$over_3sd), " (", round(sum(a$over_3sd)/nrow(a),3), "%)")
    a <- a|> filter(over_3sd == FALSE)
    
    # Make mesh
    a <- sdmTMB::add_utm_columns(a, ll_names = c("lon", "lat"), utm_crs = 3338)
    mesh <- sdmTMB::make_mesh(a, xy_cols = c("X","Y"), cutoff = cutoff)
    message("Mesh: ", mesh$mesh$n, ".")
  }
  
  # Create empty results list
  results <- list()
  
  
  # Get base model 
  st <- ifelse(obj_name == "dogfish", "off", "rw")
  fit_base_model <- function(mesh, a, fam) {
  try(sdmTMB(log(weight_kg) ~ log_length*as.factor(region) + as.factor(year) + as.factor(sex),
    mesh = mesh, spatial = "on", spatial_varying = ~ log_length,
    time = "year", spatiotemporal = st, anisotropy = TRUE,
    data = a, family = fam
  ))
  }
  
  base_path <- paste0("models/", obj_name, ".rds")
  if ("base" %in% vars) {
    message("Fitting base model...")
    base <- fit_base_model(mesh, a, fam)
  } else {
      if (file.exists(base_path)) {
      message("Loading cached base model...")
      existing_models <- readRDS(base_path)
      if (!is.null(existing_models$base)) {
        base <- existing_models$base
      } else {
        message("No 'base' entry found in cached file. Fitting base model...")
        base <- fit_base_model(mesh, a, fam)
      }
  } else {
    message("No cache found. Fitting base model...")
    fit_base_model(mesh,a, fam)
    } 
  }
  
  # Set up all model iterations
  model_iterations <- list(
    base = quote(base),
    base_centered = quote(update(base, formula = log(weight_kg) ~ log_lengthc*as.factor(region) + as.factor(year) + as.factor(sex), spatial_varying = ~ log_lengthc)),
    base_re = quote(update(base, formula. = . ~ . + (1|trawl_id))),
    temp = quote(update(base, formula. = . ~ . + s(tempc, k=4))),
    salinity = quote(update(base, formula. = . ~ . + s(salinityc, k=4))),
    temp_salinity = quote(update(base, formula. = . ~ . + s(tempc, k=4) + s(salinityc, k=4))),
    year_removal = if (obj_name != "perch") {
      quote(update(base, formula. = . ~ . - as.factor(year) + s(tempc, k=4) + s(salinityc, k=4))) 
    } else {
      quote(update(base, formula. = . ~ . - as.factor(year) + s(tempc, k=4)))
    },
    st_removal = if (obj_name != "perch") {
      quote(update(base, formula. = . ~ . + s(tempc, k=4) + s(salinityc, k=4), spatiotemporal = "off")) 
    } else {
      quote(update(base, formula. = . ~ . + s(tempc, k=4), spatiotemporal = "off"))
    },
    temp_region = quote(update(base, formula. = . ~ . + s(tempc, k=4, by = region))),
    temp_region2 = quote(update(base, formula. = . ~ . - as.factor(year) + s(tempc, k=4, by = region))),
    spline_removal = quote(update(base, formula. = . ~ . + salinityc)),
    temp_svc = quote(update(base, formula. = . ~ . + tempc, spatial_varying = ~ log_length + tempc)),
    diffusion1 = quote(update(base, nonlocal_formula = ~ diffusion(tempc))), # need to update nonlocal_data
    diffusion2 = quote(update(base, nonlocal_formula = ~ time_lag(tempc))), # need to update nonlocal_data
    dogfish0 = quote(update(base, formula = log(weight_kg) ~ log_lengthc*as.factor(region) + as.factor(year) + as.factor(sex), spatial_varying = ~ log_lengthc, spatiotemporal = "iid")), # base centered + turning st on
    dogfish1 = quote(update(base, spatiotemporal = "iid")), # base turning st on
    dogfish2 = quote(update(base, formula. = . ~ . - as.factor(year) + s(tempc, k=4))), # year removal just on temp
    dogfish3 = quote(update(base, formula. = . ~ . - as.factor(year) + s(tempc, k=4), spatiotemporal = "iid")), # above but turn st on
    dogfish4 = quote(update(base, formula. = . ~ . - as.factor(year) + tempc)), # year removal just on temp - no smooth
    dogfish42 = quote(update(base, formula. = . ~ . + tempc)),
    dogfish5 = quote(update(base, formula. = . ~ . - as.factor(year) + tempc, spatiotemporal = "iid")), # year removal on temp - no smooth, st on
    dogfish6 = quote(update(base, formula. = . ~ . - as.factor(year) + tempc, spatiotemporal = "rw")) # year removal on temp - no smooth, st rw
  )
  
  # Running model(s)
  for (model in names(model_iterations)) {
    if (!(model %in% vars)) next
    message("Modeling ", species, " ", model, "...")
    results[[model]] <- try(eval(model_iterations[[model]]))
  }
  
  # Getting the model object from env/files if no models are requested - only generate AIC table from existing list
  # if (length(vars) == 0 || length(results) == 0) {
  #   message("No models called. Pulling existing models instead from environment...")
  #   if (exists(obj_name, envir = .GlobalEnv, inherits = FALSE)) {
  #        results <- get(obj_name, envir = .GlobalEnv)
  #      } else {
  #        message("No existing models found in environment. Checking files...")
  #        if (file.exists(paste0("models/", obj_name, ".rds"))) {
  #          results <- readRDS(paste0("models/", obj_name, ".rds"))
  #        } else {
  #          message("No models found in files.")
  #          return(invisible(NULL))
  #        }
  #      }
  # } else if (length(vars) > 0) {
  #   rm(a, a1, mesh)
  #   gc()
  # }
  
  # Construct AIC tables. Pull existing tables and overwrite if necessary.
  message("Performing sanity check and constructing AIC table...")
  existing_csv <- paste0("models/", obj_name, "_aic.csv")
  prior_table <- if (file.exists(existing_csv)) read_csv(existing_csv, show_col_types = FALSE) else NULL
  check_aic(results, aic_table = prior_table, model_list_name = obj_name)
  
  # Save model list object if requested
  if (save_models == TRUE) {
    message("Saving models...")
    existing_saved <- if (file.exists(paste0("models/", obj_name, ".rds"))) {
      readRDS(paste0("models/", obj_name, ".rds"))
    } else {
      list()
    }
    existing_saved[names(results)] <- results
    assign(obj_name, existing_saved, envir = .GlobalEnv)
    saveRDS(existing_saved, paste0("models/", obj_name, ".rds"))
  }
  rm(results)
  invisible(gc())
}

# Run base models
modeler("arrowtooth flounder", "arrowtooth", vars = c("base", "base_centered", "year_removal"), dist = "student", save_models = TRUE)
modeler("pacific ocean perch", "perch", vars = c("base", "base_centered", "dogfish4"), save_models = TRUE)
# need to run again with student t dist:
modeler("shortspine thornyhead", "shortspine", vars = c("base", "base_centered", "year_removal"), dist = "student", save_models = TRUE)
modeler("dover sole", "dover", vars = c("base", "base_centered", "year_removal"), save_models = TRUE)

# reminder: the base dogfish model has spatiotemporal = "off"
modeler("pacific spiny dogfish", "dogfish", vars = c("base", "base_centered", "dogfish4"), save_models = TRUE)

# Other
modeler("pacific ocean perch", "perch", vars = c("dogfish4"), save_models = TRUE)
modeler("pacific spiny dogfish", "dogfish", vars = c("dogfish4"), save_models = TRUE)


```


# Temperature marginal effects
```{r}
# pull marginal effects of smoothing. Use sdmtmb plotting functions and use gratia library for visualizing univariate smooths, might not work with SVC but try. Other method: create prediction grid for 1 year -- with spatial effects=0, sex = female, length = avg length, single region(startwith WC), salinity = mean(salinity), temp = (make 100 vals), single X/Y coord (choose one where spatial field/omega_s ~0)
# Separate by region - WC first
# DF will have 100 rows. Pass into predict(..., se_fit = TRUE)

mod <- arrowtooth_env$year_removal
grid <- expand_grid(year = 2023, sex = "F", region = "nwfsc",
         length_cm = mean(mod$data$length_cm, na.rm=T),
         salinityc = mean(mod$data$salinityc, na.rm=T),
         X = 2470, #mean(mod$data$X,na.rm=T),
         Y = -453, #mean(mod$data$Y,na.rm=T),
         tempc = seq(from = min(mod$data$tempc), to = max(mod$data$tempc), by = 0.1))

pred <- predict(mod, grid, se_fit = TRUE)
pred <- pred|>mutate(temp = tempc + mean(mod$data$temp))

g <- ggplot(data = pred, aes(x = temp,y = est)) +
  geom_ribbon(aes(ymin = est-est_se, ymax = est+est_se), alpha=0.3) +
    geom_line() +
  theme_minimal()+
  labs(x = "Temperature (c)", title = "Arrowtooth temperature marginal effects -somewhere off WC")

ggsave("figures/test_marginal.png",g)

```

# Time series - not use since have indices
```{r}
# to do: time series for each survey for st effect, averg across year for each survey domain

models <- list(
  arrowtooth = arrowtooth$base,
  dover = dover$base,
  #dogfish = dogfish,
  perch = perch$base,
  shortspine = shortspine$base)

time_series <- purrr::map_df(models, function(mod) {
  tidy(mod, effects = "fixed")|>
    mutate(year = str_remove(term, "as.factor\\(year\\)"))
}, .id = "species")|>
  filter(str_detect(term, "year"))

ggplot(time_series, aes(x = year, y = estimate)) +
  geom_point()+
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.3) +
  facet_wrap(~species) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 0.5))
```

# Spatiotemporal maps - not use since have other maps
```{r}

preds <- list("Arrowtooth flounder" = arrowtooth_pred, 
              "Dover sole"=dover_pred, 
              "Shortspine thornyhead"=shortspine_pred,
              #"Pacific spiny dogfish" =dogfish_pred,
              "Pacific ocean perch"=perch_pred)

for (i in names(preds)) {
# Print spatial correlation facets
  pred <- preds[[i]]
  r <- max(abs(pred$epsilon_st))
  g <- ggplot() +
    stat_summary_hex(data = pred, aes(x = x, y = y, z = epsilon_st), bins = 50) +
    scale_fill_gradientn(name = "epsilon_st",
                         colours = rev(pnw_palette("Shuksan2")),
                         limits = c(-r,r)) +
    geom_sf(data = proj) +
    facet_wrap(~year) +
    theme_minimal() +
    labs(x = "", y = "", title = paste0(i, " Spatiotemporal"), color = "log(length)")
  print(g)
}
```

# Spatial / temporal ratios
```{r}
models <- list(
  arrowtooth = arrowtooth,
  dover = dover,
  dogfish = dogfish,
  perch = perch,
  shortspine = shortspine)

# Spatial / temporal ratios
ran_pars <- purrr::map_df(models, function(m) {
  tidy(m, effects = "ran_pars")
}, .id = "species")

ran_pars |>
  filter(term %in% c("sigma_O", "sigma_E")) |>
  select(species, term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate) |>
  mutate(ratio_E_O = sigma_E / sigma_O,  # spatiotemporal relative to spatial
         ratio_O_E = sigma_O / sigma_E)   # spatial relative to spatiotemporal
```

# Animations
```{r}
# g <- ggplot() +
#   stat_summary_hex(data = pred, aes(x = x, y = y, z = anomaly), bins = 50) +
#   scale_fill_gradient2(name = "spatial/temporal \nanomaly") +
#   geom_sf(data= proj) +
#   facet_wrap(~year) +
#   #geom_sf_text(data = proj, aes(label = name), size = 2.5, check_overlap = F) +
#   theme_minimal()

p <- ggplot() +
      stat_summary_hex(data = arr_pred, aes(x = x, y = y, z = est_rf), bins = 50) +
      scale_fill_gradient2(name = "random fields\nestimate")+
      geom_sf(data = proj) +
      geom_sf_text(data = proj, aes(label = name), size = 2.5, fontface = "bold", check_overlap = F) +
      theme_minimal() +
  labs(title = paste0(unique(arr_pred$species), " {closest_state}")) +
  theme(plot.title = element_text(size = 20),
        legend.title = element_text(size = 20)) +
  transition_states(year, transition_length = 1, state_length = 1)

animate(p, fps = 5, width = 800, height = 600, silent = TRUE)
```


