---
title: "Tables"
format: 
  html:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
editor: source
---
# To do:
- Aggregate regions and do species-specific KS test (or other tests) to compare length distributions
- Make tables for each species, see if there is enough data by year and survey
- Investigate depth gradients across different survyeys --> eg. nwfsc goes deeper so might have more lengths on the longer side
```{r warning = FALSE, message = FALSE,include=FALSE}
library(sdmTMB)
library(ggplot2)
library(surveyjoin)
library(rnaturalearth)
library(stringr)
library(dplyr)
library(lubridate)
library(broom)
library(tidyr)
library(sf)
library(patchwork)
library(transport)
library(kableExtra)
library(PNWColors)
library(RhpcBLASctl)
RhpcBLASctl::blas_set_num_threads(1)
RhpcBLASctl::omp_set_num_threads(1)
options(scipen = 999)
```

```{r echo = FALSE}
fish_all <- readRDS("fish_all.rds")
# Filters: no 2020 PBS fish, no unsexed fish, no entries with missing coordinates, no zeroed lengths/weights that will crash the models, no dogfish in the aleutians, no arrowtooth in the northern Bering sea
fish_all_clean <- fish_all |> 
  filter(weight_kg > 0, length_cm > 0, !is.na(lon), !is.na(temp), 
         year != 2020, sex %in% c("M", "F")) |>
  mutate(yearc = as.numeric(scale(year, center = T, scale = F)),
         tempc = temp - mean(temp),
         salinityc = salinity - mean(salinity)) |>
  mutate(type = case_when(
    is.na(weight_kg) ~ "length-only",
    !is.na(weight_kg) ~ "length + weight"
  ))
```

# Distribution tests
Wasserstein test: determine if two samples come from the same distribution. Quantifies the amount of transport you would need to get from one distribution to the other.
```{r echo =FALSE}
fish_all_clean |> filter(!survey %in% c("northern Bering Sea", "Bering Sea Slope")) |>
  filter(!common_name %in% c("sablefish","longspine thornyhead")) |>
  filter(type == "length + weight") |>
ggplot(aes(x = length_cm, color = region)) +
  stat_ecdf() +
  facet_wrap(~common_name)+
  scale_color_brewer(palette = "Dark2")+
  theme_minimal()

for (i in list("arrowtooth flounder", "dover sole", "pacific spiny dogfish", "shortspine thornyhead", "pacific ocean perch")) {
  ca1 <- fish_all_clean |> filter(region == "pbs", common_name == i)|>pull(length_cm)
  ca2 <- fish_all_clean |> filter(region == "nwfsc",common_name == i) |>pull(length_cm)
  ca3 <- fish_all_clean |> filter(region == "afsc", common_name == i) |> pull(length_cm) 
  message("Species: ", i, ". PBS sd: ", round(sd(ca1),2), ". AFSC sd: ", round(sd(ca3),2), ". NWFSC sd: ", round(sd(ca2),2))  
  message("PBS and NWFSC: ", round(wasserstein1d(ca1,ca2,p=1),2), " cm.")
  message("PBS and AFSC: ", round(wasserstein1d(ca1,ca3,p=1),2), " cm.")
  message("NWFSC and AFSC: ", round(wasserstein1d(ca2,ca3,p=1),2), " cm.")
  message("")
}
```


# Number of observations - by year and survey

## Arrowtooth flounder
- Remove northern Bering Sea (only 1 year)
- Remove Bering Sea Slope (only 3 years)
```{r echo=FALSE}
fish_all_clean |> 
  filter(common_name == "arrowtooth flounder") |>
  count(common_name,survey,year) |>kbl()
```
## Dover sole
- Remove Aleutian Islands
- Remove all Bering Sea
```{r echo=FALSE}
fish_all_clean |> 
  filter(common_name == "dover sole") |>
  count(common_name,survey,year) |>kbl()
```

## Shortspine thornyead
- Remove Bering Sea Slope (only 1 year)
- Remove all Bering Sea (no data)

- Hecate Strait no data before 2011...
- Queen Charlotte Sound no data before 2009...
- Haida Gwaii and Vancouver Island no data before 2010...
```{r echo=FALSE}
fish_all_clean |> 
  filter(common_name == "shortspine thornyhead") |>
  count(common_name,survey,year) |>kbl()
```

## Pacific ocean perch
- Bering Sea Slope (no data after 2016)
- Remove Bering Sea (no data)
```{r echo=FALSE}
fish_all_clean |> 
  filter(common_name == "pacific ocean perch") |>
  count(common_name,survey,year) |>kbl()
```

## Pacific spiny dogfish
- Remove Aleutians (only 1 year)
- Gulf of Alaska: only 4 years of substantial data, and none after 2011 --> remove?
- Hecate Strait no substantial data before 2009...
- Remove Haida Gwaii (too few observations)
```{r echo=FALSE}
fish_all_clean |> 
  filter(common_name == "pacific spiny dogfish") |>
  count(common_name,survey,year) |>kbl()
```

